---
title: "Blog Post - World Series Home Field Advantage"
output: html_document
---

```{r setup, include=FALSE}

knitr::knit_hooks$set(inline = function(x) { knitr:::format_sci(x, 'md')})
knitr::opts_chunk$set(echo = TRUE, cache = TRUE)

# This section loads necessary R libraries and sources scripts that define 
# useful functions format_md.

library(viridis)
library(ggplot2)
library(tidyverse)

```
## If home field advantage exists, how much of an impact does it have on winning the world series?


### Background

In our previous post, we looked at some discrete probability calculations to answer questions about who would win the baseball World Series given certain conditions. The World Series is a best of 7 tournament between the American and National Leagues of Major League Baseball. In this post, we will add another dimension to that problem to make it more realistic. We will add the factor of home-field advantage. 

The home field advantage is the edge which a team may have when playing a game at its home stadium. For example, it is the edge the Braves may have over the Yankees when the head-to-head match-up is in Atlanta. It is the advantage the Yankees may have when the head-to-head match-up is in New York.

In this blog post we will use simulation and analytic methods to compare the probability of winning the World Series with and without home field advantage.

First, let's take a look at the setup for this simulation. 

### Setup:

-   Suppose that the Braves and the Yankees are teams competing in the
    World Series.

-   The table below has the two possible schedules for each game of the
    series. (NYC = New York City, ATL = Atlanta)

| Overall advantage | Game 1 | Game 2 | Game 3 | Game 4 | Game 5 | Game 6 | Game 7 |
|:-----------------:|:------:|:------:|:------:|:------:|:------:|:------:|:------:|
|       Braves      |   ATL  |   ATL  |   NYC  |   NYC  |   NYC  |   ATL  |   ATL  |
|      Yankees      |   NYC  |   NYC  |   ATL  |   ATL  |   ATL  |   NYC  |   NYC  |

-   Let *P*<sub>*B*</sub> be the probability that the Braves win a
    single head-to-head match-up with the Yankees, under the assumption
    that home field advantage doesn’t exist. Let
    *P*<sub>*B*</sub><sup>*H*</sup> denote the probability that the
    Braves win a single head-to-head match-up with the Yankees as the
    home team (H for home). Let *P*<sub>*B*</sub><sup>*A*</sup> denote
    the probability that the Braves win a single head-to-head match-up
    with the away team (A for away).

| Game location |    No advantage   | Advantage                                                            |
|:-------------:|:-----------------:|:---------------------------------------------------------------------|
|      ATL      | *P*<sub>*B*</sub> | *P*<sub>*B*</sub><sup>*H*</sup> = *P*<sub>*B*</sub> \* 1.1           |
|      NYC      | *P*<sub>*B*</sub> | *P*<sub>*B*</sub><sup>*A*</sup> = 1 − (1 − *P*<sub>*B*</sub>) \* 1.1 |


### Questions:

Questions to answer:

1.  Compute analytically the probability that the Braves win the world
    series when the sequence of game locations is {NYC, NYC, ATL, ATL,
    ATL, NYC, NYC}. (The code below computes the probability for the
    alternative sequence of game locations. **Note:** The code uses
    `data.table` syntax, which may be new to you. This is intential, as
    a gentle way to introduce `data.table`.) Calculate the probability
    with and without home field advantage when *P*<sub>*B*</sub> = 0.55.
    What is the difference in probabilities?
    
    In this method of trying to compute the probability that the Braves win the wrold series given a certain sequence of game locations, we use and modify the below code chunk provided to us. First, a data frame containing all the possible world series outcomes is read in. A home field indicator is established, which indicates what the series of playing home and away games would be. Variables for the probability of winning a single game, the advantage multiplier (10% advantage) (equal to 1 for no advantage), the probability of winning a home game and an away game are created. Then we calculate the probability of each possible outcome in a series of 7 games, given the home field indicator and the advantage multiplier. Another column containing the probability of winning each possible outcome is added to the initial data frame. The sum of the probability of winning or losing each game provides the probability of winning or losing the world series.

  
```{r}
require(dplyr)
require(data.table)
# Get all possible outcomes
apo <- fread("all-possible-world-series-outcomes.csv")

# Home field indicator
hfi1 <- c(1,1,0,0,0,1,1) #{ATL, ATL, NYC, NYC, NYC, ATL, ATL}
hfi2 <- c(0,0,1,1,1,0,0) #{NYC, NYC, ATL, ATL, ATL, NYC, NYC}

# P_B
pb <- 0.55
advantage_multiplier <- 1 # Set = 1 for no advantage
pbh <- 0.55*advantage_multiplier
pba <- 1 - (1 - 0.55)*advantage_multiplier

# Calculate the probability of each possible outcome for hfi2
apo[, p := NA_real_] # Initialize new column in apo to store prob
for(i in 1:nrow(apo)){
  prob_game <- rep(1, 7)
  for(j in 1:7){
    p_win <- ifelse(hfi2[j], pbh, pba) #Change hfi depending on what series is running
    prob_game[j] <- case_when(
        apo[i,j,with=FALSE] == "W" ~ p_win
      , apo[i,j,with=FALSE] == "L" ~ 1 - p_win
      , TRUE ~ 1
    )
  }
  apo[i, p := prod(prob_game)] # Data.table syntax
}

# Sanity check: does sum(p) == 1?
#apo[, sum(p)] # This is data.table notation

# Probability of overall World Series outcomes
AWinWithoutAd <- apo[, sum(p), overall_outcome]
print(paste0("Probability of winning without the home field advantage = ", AWinWithoutAd$V1[[1]]))

# Calculate the probability of each possible outcome for hfi2 without home field advantage
# P_B
pb <- 0.55
advantage_multiplier <- 1.1 # Set = 1 for no advantage
pbh <- 0.55*advantage_multiplier
pba <- 1 - (1 - 0.55)*advantage_multiplier
apo[, p := NA_real_] # Initialize new column in apo to store prob
for(i in 1:nrow(apo)){
  prob_game <- rep(1, 7)
  for(j in 1:7){
    p_win <- ifelse(hfi2[j], pbh, pba) #Change hfi depending on what series is running
    prob_game[j] <- case_when(
        apo[i,j,with=FALSE] == "W" ~ p_win
      , apo[i,j,with=FALSE] == "L" ~ 1 - p_win
      , TRUE ~ 1
    )
  }
  apo[i, p := prod(prob_game)] # Data.table syntax
}

# Sanity check: does sum(p) == 1?
#apo[, sum(p)] # This is data.table notation

# Probability of overall World Series outcomes
AWinWithAd <- apo[, sum(p), overall_outcome]
print(paste0("Probability of winning with the home field advantage = ", AWinWithAd$V1[[1]]))
```
The probability of the Braves winning the world series given the home field advantage (More games were played at ATL than NYC) was `r AWinWithAd$V1[[1]]`, whlie the probability of them winning without the home field advantage was `r AWinWithoutAd$V1[[1]]`, given that the probability of winning any single game without the home field advantage is 0.55 and with the advantage is 0.55 plus a 10% advantage. The difference between the probabilities is `r abs(AWinWithAd$V1[[1]] - AWinWithoutAd$V1[[1]])`.


2.  Calculate the same probabilities as the previous question by
    simulation.
    
    Now, we will calculate the probabilities generated above by the process of simulation. We will use the same variables as earlier to indicate the probability of winning a single game, the advantage multiplier (10% advantage), the probability of winning a home game and an away game. W first create a function called world series that takes in 3 parameters - the home field indicator, the probability of winning a single game (Set to 0.55 as default, unless specified), and the advantage multiplier (Set to 1 as default, unless specified). This function records the number of wins and losses, and runs a 7 game simulation of the world series. Depending on the home field indicator, the probability of winning an away or home game is set, indexing through the home field indicator. It records the game outcome using the random binomial R function to generate a win or a loss, given the probability of winning that single game depending on the home field advantage. The number of wins and losses are updated, and if the Braves win or lose, the series is ended and the probability of winning is reported. 
    
    Now that we have a function that simulates a single world series, we can simulate 100,000 runs of the world series to determine the average chance of winning the world series depending on home field advantage. We sample through all the simulatiosn, recording a win or loss of the world series in a vector called brave.win. The mean of the values in the vector will give the probability of winning the world series depeding on home field advantage. 

```{r}
#default
world_series <- function(hfi, pb=0.55, advantage_multiplier=1) { 
  wins <- 0
  games <- 0
  pbh = pb*advantage_multiplier
  pba = 1 - (1-pb)*advantage_multiplier
  for (i in 1:7) {
    if (hfi[i]) {
      p_game <- pbh
    } else {
      p_game <- pba
    }
    game_outcome <- rbinom(1, 1, p_game)
    wins <- wins + game_outcome
    games <- games + 1
    if (wins == 4 | games - wins == 4)
      break
  }
  return(wins > 3) #either won or didn't
}

########
sims <- 1000000 #increase this
num.seq <-
  sample(1:sims, 1) 
brave.win <- rep(NA, sims)

for (i in 1:sims) {
    brave.win[i] <- world_series(hfi = hfi2, advantage_multiplier = 1)
}
WinWithoutAd <- mean(brave.win)
print(paste0("Probability of winning without the home field advantage = ", WinWithoutAd))

###

sims <- 1000000 #increase this
num.seq <-
  sample(1:sims, 1)  
brave.win <- rep(NA, sims)

for (i in 1:sims) {
    brave.win[i] <- world_series(hfi = hfi2, advantage_multiplier = 1.1)
}
WinWithAd <- mean(brave.win)
print(paste0("Probability of winning with the home field advantage = ", WinWithAd))
#########
#sims <- 100000 #increase this
#num.seq <-
#  sample(1:sims, 1)  #no of times you are using hfi1 or hfi2 randomly
#brave.win <- rep(NA, sims)
#for (i in 1:num.seq) {
#  brave.win[i] <- world_series(hfi = hfi1)
#}

#for (j in (num.seq + 1):sims)  {
#  brave.win[j] <- world_series(hfi = hfi2)
#}

#mean(brave.win)
```
The probability of the Braves winning the world series given the home field advantage (More games were played at ATL than NYC) was `r WinWithAd`, whlie the probability of them winning without the home field advantage was `r WinWithoutAd`, given that the probability of winning any single game without the home field advantage is 0.55 and with the advantage is 0.55 plus a 10% advantage.


3.  What is the absolute and relative error for your simulation in the
    previous question?

To calculate the absolute and relative error, we can use the formulas for error that we saw earlier. 

absolute error = |p̂−p|

relative error = |p̂−p|/p.

In the case of this scenario, let p̂ denote the probability estimated from simulation, and p as the probability estimated from the analytic solution. 

```{r}
abs_errorhfi <- abs(WinWithAd - AWinWithAd$V1[[1]])
abs_errorhfi
rel_errorhfi <- abs_errorhfi/AWinWithAd$V1[[1]]
rel_errorhfi
abs_error <- abs(WinWithoutAd - AWinWithoutAd$V1[[1]])
abs_error
rel_error <- abs_error/AWinWithoutAd$V1[[1]]
rel_error
```
The Absolute error for winning with the home field advantage was `r abs_errorhfi`. The relative error for winning with the home field advantage was `r rel_errorhfi`. 

The Absolute error for winning with the home field advantage was `r abs_error`. The relative error for winning with the home field advantage was `r rel_error`. 

As can be seen, the error between the two methods is extremely small, and both methods provide relatively similar results. 
_________

4.  **Bonus.** Does the difference in probabilites (with vs without home
    field advantage) depend on *P*<sub>*B*</sub>?
    
```{r}

```

5.  **Bonus.** Does the difference in probabilites (with vs without home
    field advantage) depend on the advantage factor? (The advantage
    factor in *P*<sub>*B*</sub><sup>*H*</sup> and
    *P*<sub>*B*</sub><sup>*A*</sup> is the 1.1 multiplier that results
    in a 10% increase for the home team.)
    
```{r}

```
    
    