---
title: "Blog Post - Coverage Probability"
output: html_notebook
---

```{r load_packages, message=FALSE, warning=FALSE}
library(tidyverse)
library(foreach)
library(dplyr)
library(doParallel)
```

# Introduction

Coverage probability is an important operating characteristic of methods
for constructing interval estimates, particularly confidence intervals.

**Definition:** For the purposes of this blog post, we define the 95%
confidence interval of the mean to be the middle 95% of sampling
distribution of the mean. Similarly, the 95% confidence interval of the
median, standard deviation, etc. is the middle 95% of the respective
sampling distribution.

**Definition:** For the purposes of this blog post, we define the
coverage probability as the long run proportion of intervals that
capture the population parameter of interest. Conceptualy, one can
calculate the coverage probability with the following steps

1.  generate a sample of size *N* from a known distribution
2.  construct a confidence interval
3.  determine if the confidence captures the population parameter
4.  Repeat steps (a) - (c) many times. Estimate the coverage probability
    as the proportion of samples in which the confidence interval
    captured the population parameter.

The figure below shows the 95% confidence interval calculated for a
handful of samples. Intervals in blue capture the population parameter
of interest; intervals in red do not.

![](./coverage-prob.svg)

Idealy, a 95% confidence interval will capture the population parameter
of interest in 95% of samples.

# Setup

In this blog post, we will perform a simulation to calculate the
coverage probability of the 95% confidence interval of the median when
computed from *F̂*<sub>*X*</sub><sup>*m**l**e*</sup>.

We will take the following approach in trying to determine the coverage probability of the 95% confidence interval of the median. 

**Step 1:** Generate a single sample from a standard normal distribution
of size *N* = 200. In this setup, we will use MLE to estimate the distribution.

**Step 2:** Approximate the sampling distribution of the median, conditional on the estimate of the distribution generated in the previous step.

**Step 3:** Calculate a 95% confidence interval from the
approximated sampling distribution.

**Step 4:** Calculate the coverage probability.

**Step 5:** Perform the simulation and report the results.

**Step 6:** Describe how you might change the simulation to learn more
about the operating characteristics of your chosen method for
constructing the 95% confidence interval.

## Generating a single sample

```{r parameters}

parameters <- list(n = 201, dist = "rnorm", mean = 0, sd = 1)

```

```{r generate_data}
generate_data <- function(parameter) {
  if (parameter$dist == "rnorm") {
     parameter$data <- rnorm(parameter$n, parameter$mean, parameter$sd) 
  }
  return(parameter)
}

parameters <- generate_data(parameters)
```

In the first code chunk above, I create a list called parameters that I will pass to the generate_data function. This contains the sample size 'n', the type of distribution 'dist', a value for the mean, and a value for the standard deviation 'sd'. 

In the generate_data function, I check which distribution I am creating a sample for, and generate a standard normal distribution using the rnorm function and the parameters being passed in. I then add this data back into the list in a sub-list called data. 

## Estimating the MLE

```{r est.mle}
est.mle <- function(parameter) {
  data <- parameter$data
  parameter$mean_mle <- mean(data)
  parameter$sd_mle <- var(data)
  return(parameter)
}

parameters <- est.mle(parameters)
```

To estimate the MLE, I update the mean and standard deviation using the mean and standard deviation functions in R to capture the real mean and standard deviation of the data generated using the generate_data function. These values are stored in elements called mean_mle and sd_mle and are stored within the parameter list being passed through these functions. 

## Constructing the Confidence Intervals

```{r boot_meds_ci}
boot.meds.ci <- function(parameter) {
  R <- 5000 
  sample.meds <- NA
  for (i in 1:R) {
      sample.meds[i] <- parameter %>% with(rnorm(n = n, mean = mean_mle, sd = sd_mle) %>% median)
  }
  parameter$CI <- quantile(sample.meds, c(0.025, 0.975))
  return(parameter)
}

parameters <- boot.meds.ci(parameters)
```

Using the bootstrap method, I sample 5000 medians of the data created using the rnorm function with the updated mean and sd from the MLE function, and use the quantile function to estimate the middle 95% confidence interval for the median. This confidence interval is then stored in the parameter list in an element called CI. 

## Capture Medians/Summary Statistic

```{r capture_median}
capture_median <- function(parameter){
  # Hard coded 0 as parameter of interest
  if(parameter$CI[1] < 0 & 0 < parameter$CI[2]) {
    count = 1
  } else {
    count = 0
  }
  return(count)
}
capture_median(parameters)

parameters %>% generate_data %>% est.mle %>% boot.meds.ci %>% capture_median
```

I then determine whether the generated confidence interval falls within the required range surrounding 0. If it is within the range, I return a variable count with the value 1, otherwise I return a 0. 

To simulate the run of these above functions to determine the average coverage probability, I use parallel processing to run the pipeline of parameters %>% generate_data %>% est.mle %>% boot.meds.ci %>% capture_median 5000 times. I store the result of capture_median in a vector called captures. The mean of the vector captures will give the coverage probability. 

```{r parallel_run}
parameters <- list(n = 201, dist = "rnorm", mean = 0, sd = 1)
require(foreach)
require(dplyr)
# Create parallel backend
require(doParallel)
cores_2_use <- detectCores() -1
# Create parallel backend
cl <- makeCluster(cores_2_use)
clusterSetRNGStream(cl, 2344)
registerDoParallel(cl)

captures <- foreach(
    i = 1:5000
  , .combine = c
  , .packages = c('dplyr') # Need to indicate which packages
) %dopar% {
  parameters %>% generate_data %>% est.mle %>% boot.meds.ci %>% capture_median
}
Coverage_Probability <- mean(captures)
stopCluster(cl)
Coverage_Probability
```

As can be seen from the result of the simulation above, there is about a 98.34% coverage probability of the 95% confidence interval of the median when computed from *F̂*<sub>*X*</sub><sup>*m**l**e*</sup>.
